\chapter{Coulomb-Sturmian-based quantum chemistry}
\label{ch:CSQChem}
\chapquote{%
The real problem is that programmers have spent far too much
time worrying about efficiency in the wrong places and at the
wrong times; premature optimization is the root of all evil
(or at least most of it) in programming.}
{Donald Knuth~(1928--present)}
\nomenclature{$\kopt$}{Optimal Coulomb-Sturmian exponent $\kexp$,
	which yields the best description of the system
	given the current Coulomb-Sturmian basis and
	the selected quantum-chemical method.}
\nomenclature{$\RMSOl$}{
	The root mean square occupied coefficient per angular momentum $l$,
	see definition \vref{defn:RMSOl}
}

% TODO Not sure this guy is needed
\newcommand{\limp}{l_\text{imp}}

\noindent
In this chapter we will discuss a preliminary computational results
obtained from quantum-chemical calculations using Coulomb-Sturmian
basis functions inside the \molsturm framework.
The focus will be on discussing and understanding the convergence
properties of \CS-based calculations on atoms of
the second and the third period of the periodic table,
mostly at \HF or {\MP}2 level.
Some exemplary Full-CI calculations have been performed
to obtain reliable reference results
and to show that the convergence trends of {\MP}2 and Full-CI agree.

Even though these methods represent only a small selection
of the wide-range of possible quantum-chemical methods,
we think that Hartree-Fock as a self-consistent field method
and {\MP}2 to perturbatively capture correlation effects
provide good insight into what could be expected
for other single-reference ground state methods
if Coulomb-Sturmian basis sets are to be used.

Based on these results we therefore suggest
preliminary guidelines for selecting \CS basis set parameters,
like the angular momentum restrictions or the Sturmian exponent $\kexp$,
where the overall aim is to yield rapid convergence
of the ground state energies at \HF or correlated level.
Finally we present the first results of a \CS-based
excited states calculation at \ADC-(2) level.

\input{denoting_cs_basis.tex}
\input{convergence_hf.tex}
\input{convergence_correlation.tex}
\input{convergence_kexp.tex}
\input{suggestions_basis_sets.tex}
\input{sturmian_cgto.tex}
\input{sturmian_adc.tex}
