\chapter*{Publications}
\phantomsection
\addcontentsline{toc}{chapter}{Publications}
\markboth{\leftheadstyle{Publications}}{}

\section*{Articles in preparation}
\phantomsection
\addcontentsline{toc}{section}{Articles in preparation}
\begin{itemize}
	\item M. F. Herbst, A. Dreuw and J. E. Avery.
		\textit{\texttt{molsturm}: A light-weight quantum-chemistry framework for rapid method development not restricted to a particular type of basis function}. In preparation.
	\item M. F. Herbst and A. Dreuw.
		\textit{Judging the angular momentum requirements for the representation of a wave function from Coulomb-Sturmian convergence progressions}. In preparation.
	\item M. F. Herbst, J. E. Avery and A. Dreuw.
		\textit{Coulomb-Sturmian based excited electronic states using the algebraic diagrammatic construction scheme of the polarisation propagator}. \linebreak In preparation.
	\item M. F. Herbst, M. R. B. Kristensen, J. E. Avery, B. Vinter, A. Dreuw.
		\textit{Lazyten and bohrium: Combining efforts for an automatically parallelising readable code}. \linebreak In preparation.
%	\item M. F. Herbst, J. E. Avery and A. Dreuw.
%		\textit{A systematic comparison of contracted Gaussian and Coulomb-Sturmian basis sets for electronic structure calculations}. \linebreak In preparation
\end{itemize}


\section*{Scientific software}
\phantomsection
\addcontentsline{toc}{section}{Scientific software}
This section gives the list of scientific software projects,
which I started in collaboration with others
and which are released or will be released soon.

\paragraph*{\molsturm modular and flexible quantum chemistry package}
\begin{itemize}
	\item M. F. Herbst and J. E. Avery.
		\textit{\molsturm 0.0.2}. \url{https://molsturm.org} (2017)
	\item M. F. Herbst and J. E. Avery.
		\textit{\gscf 0.1.0}. \url{https://molsturm.org/gscf} (2017)
	\item M. F. Herbst and J. E. Avery.
		\textit{\gint 0.0.0}. \url{https://molsturm.org/gint} (2017)
	\item M. F. Herbst and J. E. Avery.
		\textit{\lazyten 0.4.1}. \url{https://lazyten.org} (2017)
	\item M. F. Herbst and J. E. Avery.
		\textit{\krims 0.2.1}. \url{https://lazyten.org/krims} (2017)
	\item J. E. Avery and M. F. Herbst.
		\textit{\sturmint 0.1.0}. To be realeased.
\end{itemize}

\pagebreak
\paragraph*{\texttt{look4bas}: Interactive tool to search for Gaussian basis sets online}
\begin{itemize}
	\item M. F. Herbst. \textit{\texttt{look4bas}: First release}. (2018) DOI \mbox{10.5281/zenodo.1188992}
\end{itemize}

\section*{Lecture notes}
\phantomsection
\addcontentsline{toc}{section}{Lecture notes}
The following lecture notes were written and published during my graduate studies.
\begin{itemize}
	\item M. F. Herbst. \textit{Advanced Bash Scripting 2015}. (2015) \\
		DOI \mbox{10.5281/zenodo.1038526}
	\item M. F. Herbst. \textit{Introduction to awk programming 2016}. (2016) \\
		DOI \mbox{10.5281/zenodo.1038522}
	\item M. F. Herbst. \textit{Advanced Bash Scripting 2017}. (2017) \\
		DOI \mbox{10.5281/zenodo.1045332}
\end{itemize}

