\section{Full configuration interaction}
\label{sec:FCI}
\nomenclature{$\Phi$}{Slater determinant $\bigwedge_{i=1}^{\Nelec} \psi_i$
or many-electron basis functions}
\nomenclature{$\psi_i$}{One-particle function,
typically $i$-th eigenfunction of the Fock operator, i.e.~a Hartree-Fock orbital}
\nomenclature{$\Ibas$}{Index set of the one-particle basis functions.
	Typically a set of multi-indices of quantum numbers.}
\nomenclature{$\Nbas$}{Cardinality of $\Ibas$, i.e.~the number of one-particle basis functions.}
\nomenclature{$\varphi_\mu$}{$\mu$-th one-particle basis function of the one-particle basis $\{\varphi_\mu\}_{\mu \in \Ibas}$}

\nomenclature{$\Iocc$}{Index set of occupied SCF orbitals.}
\nomenclature{$\Ivirt$}{Index set of virtual, i.e. unoccupied, SCF orbitals.}

In this section we want to develop a numerical treatment
for solving the electronic Schrödinger equation \eqref{eqn:ElectronicSchrödinger}
under the Ritz-Galerkin projection ansatz of section \vref{sec:RitzGalerkin}.
In the previous section we analysed the mathematical implications
of the spin statistics theorem for electrons as fermionic systems,
which lead us to choose the form domain
\[ Q(\Op{H}_{\Nelec}) = H^1(\R^{3\Nelec}, \C) \cap \bigwedge^{\Nelec} L^2(\R^3, \C). \]
for the electronic Schrödinger operator $\Op{H}_{\Nelec}$.


For simplifying our treatment
we will not try to discretise this domain in the Ritz-Galerkin ansatz
of definition \ref{defn:RitzGalerkin},
much rather we will develop methods to sample only the subspace
\[ \tilde{Q}(\Op{H}_{\Nelec}) = \bigwedge^{\Nelec} H^1(\R^3, \C) \subset Q(\Op{H}_{\Nelec}) \]
due to its simpler structure.
Since this subspace is dense this implies no potential loss of numerical accuracy.
By definition of the exterior power
\begin{equation}
	\tilde{Q}(\Op{H}_{\Nelec}) = \spacespan \left\{
		\bigwedge_{i=1}^{\Nelec} \psi_i
		\, \middle| \,
		\psi_i \in H^1(\R^3, \C) \, \forall i = 1, \ldots, \Nelec
	\right\}.
	\label{eqn:FormDomainAllFunctions}
\end{equation}
Since $H^1(\R^{3\Nelec}, \C)$ is separable, we can find a countable basis set
\begin{equation}
	\set{B}_1 \equiv \{\psi_i\}_{i \in \N} \qquad \text{with }
	\braket{\psi_i}{\psi_j}_1 = \delta_{ij}
	\quad \text{and} \quad \spacespan \set{B}_1 = H^1(\R^{3\Nelec}, \C),
	\label{eqn:OneParticleBasis}
\end{equation}
where we used the abbreviated notation
$\braket{\slot}{\slot}_1 = \braket{\slot}{\slot}_{L^2(\R^3, \C)}$.
$\set{B}_1$ is sometimes called the \newterm{one-particle basis}.
Taking the properties of the wedge product \eqref{eqn:PropertiesExteriorProduct}
into account allows to deduce the equivalent construction
\begin{equation}
	\label{eqn:FormDomainSlaterDeterminants}
	\tilde{Q}(\Op{H}_{\Nelec}) = \spacespan
	\left\{ \bigwedge_{i=1}^{\Nelec} \psi_i
	, \middle| \, \psi_i \in \set{B}_1 \, \forall i=1,\ldots,\Nelec
	\right\},
\end{equation}
which builds $\tilde{Q}(\Op{H}_{\Nelec})$ as the span
over all Slater determinants built by selecting
$\Nelec$ functions from $\set{B}_1$.
Since nothing stops us from selecting the same basis function
twice from $\set{B}_1$ in this construction,
many of the constructed determinants $\bigwedge_{i=1}^{\Nelec} \psi_i$
are zero.
In other words these determinants amount to span $\tilde{Q}(\Op{H}_{\Nelec})$,
but they are not a basis for this space.
In the following we want to fix this and construct an orthonormal basis
of suitable Slater determinants.
This requires an appropriate inner product.
\begin{defn}
	Let $\tilde{Q}(\Op{H}_{\Nelec})$ be defined as in \eqref{eqn:FormDomainAllFunctions}.
	We define an inner product on $\tilde{Q}(\Op{H}_{\Nelec})$ by
	requiring for any two arbitrary Slater determinants
	\begin{align*}
		\Psi &= \bigwedge_{i=1}^{\Nelec} \psi_i
		&&\text{and}& \Xi &= \bigwedge_{i=1}^{\Nelec} \xi_i
	\end{align*}
	with $\psi_i, \xi_i \in H^1(\R^3, \C)$ for all $i \in 1,\ldots,\Nelec$:
	\begin{align}
		\braket{\Psi}{\Xi}_{\Nelec} \equiv \det \mat{G}
		\qquad
		\text{where } G_{ij} = \braket{\psi_i}{\xi_j}_{L^2(\R^3, \C)} \forall i,j \in 1,\ldots,\Nelec.
		\label{eqn:InnerProductFormDomain}
	\end{align}
	The inner product for other elements from $\tilde{Q}(\Op{H}_{\Nelec})$
	is then constructed in accordance
	with the axioms shown in definition \vref{def:InnerProduct}.
	\todoil{A more formal proof would be nice,
		especially for the completeness under this inner product.
		Or showing equivalence with respect to the inner product
		on $H^1(\R^{3\Nelec}, \C$ for the elements of $\tilde{Q}(\Op{H}_{\Nelec})$
		would be nice.}
\end{defn}

\noindent
With this inner product at hand we can construct an orthonormal basis for
$\tilde{Q}(\Op{H}_{\Nelec})$.

\begin{rem}[Orthonormal basis for $\tilde{Q}(\Op{H}_{\Nelec})$]
	\label{rem:Determinants}
	Let $\tilde{Q}(\Op{H}_{\Nelec})$ be defined as in \eqref{eqn:FormDomainAllFunctions}
	and let $\set{B}_1$ be an arbitrary one-particle basis for $H^1(\R^3, \C)$.
	We take one arbitrary, non-trivial Slater determinant
	$0 \neq \Phi_0 \in \tilde{Q}(\Op{H}_{\Nelec})$,
	such that
	\[
		\Phi_0 = \varphi_1 \wedge \varphi_2 \wedge \cdots \varphi_i \cdots \wedge \varphi_{\Nelec}.
	\]
	This determinant can always be found due to the alternative
	construction for $\tilde{Q}(\Op{H}_{\Nelec})$ sketched
	in \eqref{eqn:FormDomainSlaterDeterminants}.
	Let us call $\Phi_0$ the \newterm{reference determinant}
	in the following.

	The functions of the (countable) basis set $\{\psi_i\}_{i \in \N} = \set{B}$
	can be indexed in such a way
	that the first $\Nelec$ functions coincide with $\{\varphi\}_{i \in \Nelec}$.
	In other words
	\[
		\Phi_0 = \psi_1 \wedge \psi_2 \wedge \cdots \psi_i \cdots \wedge \psi_{\Nelec}
	\]
	as well.
	We further define the index sets%
	\footnote{%
		The subscript ``occ'' stands for \textit{occupied}
		and ``virt'' for \textit{virtual}.
		These terms will become clear when we discuss the Hartree-Fock ansatz
		in the next section.
	}
	\begin{align*}
		\Iocc &= \{1, \ldots, \Nelec\}
		&&\text{and} &
		\Ivirt &= \{ i \in \N \, | \, i > \Nelec \}.
	\end{align*}
	With reference to $\Phi_0$ we can construct
	for each $i \in \Iocc$ and each $a \in \Ivirt$
	a so-called singly \newterm{excited determinant}
	\[
		\Phi_i^a = \psi_1 \wedge \psi_2 \wedge \cdots \psi_a \cdots \wedge \psi_{\Nelec}
	\]
	by replacing the $i$-th function of the Slater determinant
	wedge string
	by the $a$-th function of $\set{B}$
	without changing the order.
	Analogously one may define doubly or higher excited determinants
	\begin{align*}
		\Phi_{ij}^{ab} &= \psi_1 \wedge \psi_2 \wedge \cdots \wedge \psi_a \cdots \psi_b
			\wedge \cdots \wedge \psi_{\Nelec} \\
		\Phi_{ijk}^{abc} &= \psi_1 \wedge \psi_2 \wedge \cdots \wedge \psi_a \cdots \psi_b
			\cdots \psi_c \wedge \cdots \wedge \psi_{\Nelec}
	\end{align*}
	where%
	\footnote{%
		This is the typical indexing convention in quantum chemistry.
		Indices $i,j,k,l,m, \ldots$ stand for occupied indices
		and $a,b,c,d,e, \ldots$ for virtual indices.
	} $i,j,k \in \Iocc$ and $a,b,c \in \Ivirt$
	In this case one has to additionally
	require that $i < j < k < \cdots$ and $a < b < c < \cdots$,
	because otherwise no new determinants are generated (if $i=j$ or $i=k$ or \ldots)
	or a zero determinant is generated (if $a=b$ or similar).
	Constructed in this way all determinants in the set
	\[
		\set{B}_{\Nelec} \equiv
		\left\{
			\Phi_0, \Phi_i^a, \Phi_{ij}^{ab}, \Phi_{ijk}^{abc}, \cdots \right\}
	\]
	are unique.
	Still it is not hard to see
	that $\spacespan \set{B}_{\Nelec} = \tilde{Q}(\Op{H}_{\Nelec})$,
	since we only took away those determinants adding redundant information
	in the construction \eqref{eqn:FormDomainSlaterDeterminants}.

	With the inner product defined in \eqref{eqn:InnerProductFormDomain} we notice
	for all $r,s \in \N$
	\[
		\braket{\Phi_0}{\Phi_r^s}_{\Nelec}
			= \braket{\psi_r}{\psi_s}_1 = \delta_{rs},
	\]
	since by \eqref{eqn:OneParticleBasis}
	all functions in $\set{B}_1$ are orthonormal to each other.
	In other words $\set{B}_{\Nelec}$ is an orthonormal
	basis for $\tilde{Q}(\Op{H}_{\Nelec})$.
	\todo[inline, caption={}]{
		This following statement might not be true,
		since $\braket{\slot}{\slot}_{\Nelec}$
		might not be equivalent to the Hilbertian inner product!
		\begin{center}
		Since it is countable, $\tilde{Q}(\Op{H}_{\Nelec})$ is separable.
		\end{center}
	}
\end{rem}
The set $\set{B}_{\Nelec}$ is sometimes called the \newterm{$\Nelec$-particle basis}
or \newterm{many-particle basis}
corresponding to $\set{B}_1$ and the reference determinant $\Phi_0$.
Notice that any choice of $\Phi_0$ or $\set{B}_1$ would work for its construction
and albeit the precise entries in $\set{B}_{\Nelec}$ might differ from case to case
the end result $\spacespan \set{B}_{\Nelec} = \tilde{Q}(\Op{H}_{\Nelec})$
is always true.

\begin{rem}
	Given a many-particle basis $\set{B}_{\Nelec}$ consisting
	of normalised Slater determinants,
	any function $\Psi \in \tilde{Q}(\Op{H}_{\Nelec})$
	can be expanded as such
	\begin{align}
		\label{eqn:ExpansionSlaterDeterminant}
		\Psi &= \sum_\mu c_\mu \Phi_\mu
		&\text{where } \forall \mu \in \N: \quad
		\Phi_\mu \in \set{B}_{\Nelec}, c_\mu \in \C.
	\end{align}
	If one is interested in emphasising the particular
	one-particle basis $\set{B}_1$ and the particular reference determinants $\Phi_0$
	this can be written equivalently as
	\begin{equation}
		\label{eqn:CIExpansion}
		\Psi
		= c_0 \Phi_0 + \sum_{ia} c_i^a \Phi_i^a
		+ \sum_{\substack{i<j \\a<b}} c_{ij}^{ab} \Phi_{ij}^{ab}
		+ \sum_{\substack{i<j<k \\ a<b<c}} c_{ijk}^{abc} \Phi_{ijk}^{abc}
		 + \cdots,
	\end{equation}
	where $i,j,k, \ldots \in \Iocc$ and $a,b,c, \ldots \in \Ivirt$.

	This expansion is commonly referred to as the \newterm{CI expansion},
	where CI stands for configuration interaction,
	a term which will become more clear
	after the next remark.
\end{rem}

\defineabbr{FCI}{FCI\xspace}{Full configuration interaction,
	see section \vref{sec:FCI}}
\newcommand{\Nfci}{N_\text{FCI}}
\begin{rem}[Full CI]
	\label{rem:FCI}
	The discrete formulation of the Ritz-Galerkin scheme
	of remark \vref{rem:DiscreteFormulation}
	can now be applied rather easily to the electronic Schrödinger equation.
	This leads to a procedure called \newterm{full CI}
	or full configuration interaction~(\FCI).

	\begin{itemize}
	\item Take a finite-sized one-electron basis set
	\[ \set{B}_1^{(n)} \equiv \{\varphi_\mu\}_{\mu \in \Ibas} \]
	of orthonormal functions $\varphi_\mu \in U$,
	where $U \subset H^1(\R^3, \C)$ is dense.
	%
	\item Define a reference determinant
	\[ \Phi_0 = \varphi_1 \wedge \varphi_2 \wedge \ldots \wedge \varphi_{\Nelec} \]
	where $\{\varphi_i\}_{i=1,\ldots,\Nelec} \subset \{\varphi_\mu\}_{\mu \in \Ibas}$
	arbitrarily and construct the finite $\Nelec$-electron basis
	\[ \set{B}_{\Nelec}^{(n)} \equiv \{ \Phi_0, \Phi_i^a, \Phi_{ij}^{ab}, \ldots \} \]
	according to the procedure described in remark \vref{rem:Determinants}.
	As usual we take
	\begin{align}
		\label{eqn:CondOccIndex}
		i, j, k, l, \ldots &\in \Iocc &&\text{with}& i&<j<k<l<\cdots \\
		\label{eqn:CondVirtIndex}
		a, b, c, d, \ldots &\in \Ivirt &&\text{with}& a&<b<c<d<\cdots
	\end{align}
	where in the finite case
	\begin{align*}
		\Iocc &= \{1, \ldots, \Nelec\} &&\text{and}& \Ivirt &= \{ \Nelec + 1, \ldots, \Nbas \}.
	\end{align*}
	%
	\item Construct the full CI matrix $\mat{A}_\text{FCI} \in \C^{\Nfci \times \Nfci}$
		consisting of elements
		\begin{equation}
			a(\Phi_1, \Phi_2)
			= \mbra{\Phi_1}
			\Op{T}_e + \Op{V}_{ne} + \Op{V}_{ee}
			\mket{\Phi_2}_{\Nelec}
			\label{eqn:FCIMatrixElements}
		\end{equation}
		for all combinations $\Phi_1, \Phi_2 \in \set{B}_{\Nelec}^{(n)}$.
		There are
		\[ \Nfci = \binom{\Nbas}{\Nelec} \leq \Nbas^{\Nelec} \]
		such Slater determinants.
	\item Diagonalise $\mat{A}_\text{FCI}$ to find a few energy eigenvalues
		$E_i^{(n)} \in \R$ and corresponding \newterm{CI vectors}
		$\vec{c}^{(n)}_i \in \C^{\Nfci}$.
	\item Repeat with a larger basis $\set{B}_1^{(n+1)}$ until convergence.
	\end{itemize}
	Notice that the subspace sequence
	$\spacespan \set{B}_{\Nelec}^{(n)} \subset \tilde{Q}(\Op{H}_{\Nelec})$
	satisfies the required condition \eqref{eqn:CondSubspaces}
	since $U$ is dense in $H^1(\R^3, \C)$,
	which makes $\spacespan \set{B}_{\Nelec}^{(n)}$
	dense in $\tilde{Q}(\Op{H}_{\Nelec})$
	and thus transitively dense in $Q(\Op{H}_{\Nelec})$.
	If $\Op{H}_{\Nelec}$ thus has a discrete ground and some discrete
	excited states below the essential spectrum,
	we can approximate it by this procedure up to arbitrary accuracy.
	This is satisfied for all neutral or positively charged systems
	and some negatively charged systems.
	Recall remark \vref{rem:ElectronicTISENumerical} for details.
\end{rem}
The term configuration interaction becomes more clear taking
a look at \eqref{eqn:FCIMatrixElements}.
Essentially a single Slater determinant describes a possible arrangement
or configuration of $\Nelec$ electrons inside the single-particle functions
from the single-particle basis $\{\varphi_\mu\}_{\mu\in\Ibas}$.
The full CI matrix of \eqref{eqn:FCIMatrixElements} describes their
physical interaction in the sense of the electronic Hamiltonian $\Op{H}_{\Nelec}$.
By diagonalising the matrix $\mat{A}_\text{FCI}$ we thus determine
the electronic eigenstates taking the full range of interactions between
all single-particle configurations into account,
hence the name full CI.
In contrast to this are the class of truncated schemes,
which do not consider all determinants for building the
approximated many-electron basis $\set{B}^{(n)}_{\Nelec}$.
We will discuss these in section \vref{sec:Correlation}.

\todoil{Rephrase from here til end of the section}
It becomes especially apparent why one wants
to use a truncated scheme if one considers the size of the
full CI matrix.
Even for a relatively small problem like water with $\Nelec = 10$
the full CI matrix is comparatively large.
For this reason in many cases one does not systematically
converge the full CI solution, but just performs the calculation once.
As we will see in the next remark the full CI matrix is not dense,
but has a banded structure, whereas the width of the band is rather
large, too.

We will conclude this chapter about full CI
with a discussion of the structure of the full CI matrix.

\defineabbr{ERI}{ERI\xspace}{Electron repulsion integrals}
\newcommand{\eriAsym}[2]{\left\langle #1 \middle|\middle| #2 \right\rangle}
\newcommand{\eriPh}[2]{\left\langle #1 \middle| #2 \right\rangle}
\newcommand{\eriMu}[2]{\left( #1 \middle| #2 \right)}
\nomenclature{$(ij|kl)$}{%
	Electron repulsion integrals in chemist's or Mullikan notation,
	see definition in remark \vref{rem:ERI}.
}
\begin{rem}[Structure and evaluation of the FCI matrix]
	Recall the structure of the electronic Hamiltonian as
	\[
		\Op{H}_{\Nelec}
		= \Op{T}_e + \Op{V}_{en} + \Op{V}_{ee}
		= \sum_{i=1}^{\Nelec}
		  \left( -\frac12 \Delta_{\vec{r}_i} + \sum_{A=1}^M \frac{Z_A}{r_{iA}} \right)
		+ \sum_{i=1}^{\Nelec} \sum_{j=i+1}^{\Nelec} \frac{1}{r_{ij}}
	\]
	and let us define the \newterm{core Hamiltonian}
	\begin{equation}
		\Op{H}_\text{core}
		= \Op{T} + \Op{V}_0
		= -\frac12 \Delta + \sum_{A=1}^M \frac{Z_A}{\norm{\vec{r} - \vec{r}_A}_2}
		\label{eqn:HCore}
	\end{equation}
	consisting of kinetic energy and nuclear attraction.

	The operators $\Op{T}_e$, $\Op{V}_{ne}$ and the
	electronic-electronic Coulombic interaction potential
	$\Op{V}_{ee}$
	differ in a very fundamental property,
	namely the number of electronic coordinates
	they make reference to.
	To distinguish this one calls operators like $\Op{T}_e$
	and $\Op{V}_{ne}$ \newterm{one-electron operator}
	and $\Op{V}_{ee}$ is an example of a \newterm{two-electron operator}.

	\todoil{Derivation of Slater-Condon rules in appendix?}
	The Slater-Condon rules~\cite{Szabo1996} now state for a single particle operator
	like the core Hamiltonian
	we have for all $a,b,c$ and $i,j,k$ under the usual conditions
	\eqref{eqn:CondOccIndex} and \eqref{eqn:CondVirtIndex}
	\begin{align*}
		\braket{\Phi_0}{\left(\Op{T}_e + \Op{V}_{en}\right) \,\Phi_0}_{\Nelec}
		&= \sum_{i\in\Iocc} \braket{\psi_i}{\Op{H}_{\text{core}} \, \psi_i}_1 \\
		%
		\braket{\Phi_0}{\left(\Op{T}_e + \Op{V}_{en}\right) \,\Phi_{i}^{a}}_{\Nelec}
		&= \braket{\psi_i}{\Op{H}_\text{core} \, \psi_a}_1 \\
		%
		\braket{\Phi_0}{\left(\Op{T}_e + \Op{V}_{en}\right) \,\Phi_{ij}^{ab}}_{\Nelec}
		&= 0.
	\end{align*}
	For the two-electron operators like $\Op{V}_{ee}$, however,
	\begin{align*}
		\braket{\Phi_0}{\Op{V}_{ee}\, \Phi_0}_{\Nelec}
		&= \frac12 \sum_{i\in\Iocc}
			  \eriMu{\psi_i\psi_i}{\psi_j\psi_j}
			- \eriMu{\psi_i\psi_j}{\psi_i\psi_j} \\
		%
		\braket{\Phi_0}{\Op{V}_{ee}\, \Phi_i^a}_{\Nelec}
		&= \sum_{j\in\Iocc}
			  \eriMu{\psi_i\psi_a}{\psi_j\psi_j}
			- \eriMu{\psi_j\psi_a}{\psi_i\psi_j} \\
		%
		\braket{\Phi_0}{\Op{V}_{ee}\, \Phi_{ij}^{ab}}_{\Nelec}
		&= \eriMu{\psi_i\psi_j}{\psi_a\psi_b}
		 - \eriMu{\psi_a\psi_j}{\psi_i\psi_b} \\
		%
		\braket{\Phi_0}{\Op{V}_{ee}\, \Phi_{ijk}^{abc}}_{\Nelec} &= 0.
	\end{align*}
	where we used the \newterm{electron repulsion integrals}~ERI
	in Mulliken convention
	\begin{equation}
		\eriMu{\psi_i \psi_j}{\psi_k \psi_l}
			= \int_{\R^3} \int_{\R^3}
				\frac{\cc{\psi_i}(\vec{r}_1) \psi_j(\vec{r}_1)
					\,\cc{\psi_k}(\vec{r}_2) \psi_l(\vec{r}_2)}
				{\norm{\vec{r}_1 - \vec{r}_2}_2}
				\D \vec{r}_1 \D \vec{r}_2
		\label{eqn:ERI}
	\end{equation}

	Higher excitations are zero as well which implies
	that $\mat{A}_\text{FCI}$ is rather sparse.

	\todoil{Figure of FCI matrix structure?}
\end{rem}

\begin{rem}[Formulation of the repulsion integrals]
	\label{rem:ERI}
	Multiple conventions exist for the repulsion integrals.
	In \eqref{eqn:ERI} we introduced the shell pair notation
	or Chemists' notation.
	Alternatively there is the physicists' convention
	\[
		\eriPh{\psi_i \psi_j}{\psi_k \psi_l}
		\equiv \eriMu{\psi_i\psi_k}{\psi_j\psi_l}
	\]
	With this we can define an antisymmetrised electron repulsion
	integral
	\[ \eriAsym{\psi_i \psi_j}{\psi_k \psi_l}
		\equiv \eriPh{\psi_i \psi_j}{\psi_k \psi_l}
		- \eriPh{\psi_j \psi_i}{\psi_k \psi_l}
		= \eriMu{\psi_i\psi_k}{\psi_j\psi_l} - \eriMu{\psi_j\psi_k}{\psi_i\psi_l}
	\]
	if the one-particle basis and its indexing is clear from context
	one often just writes
	\[ \eriMu{ij}{kl}, \eriPh{ij}{kl}, \eriAsym{ij}{kl} \]
\end{rem}
