\section{Correspondence of classical and quantum mechanics}
\label{sec:QMCorrespondence}
% TODO OPTIONAL
%\to doil{One could add a few more references for the general QM introduction part,
%	\eg to the original Schrödinger paper
\newcommand{\clphase}{(q_1, \ldots, q_d, p_1, \ldots p_d)}
%
%TODO OPTIONAL \to doil{Hint at other formulations of classical mechanics}
According to the Hamiltonian formulation of classical mechanics
a physical system with $d$ degrees of freedom
is described by a set of generalised coordinates
$q_1, \ldots, q_d$ along with their canonical momenta $p_1, \ldots, p_d$.
It is assumed that any physically measurable quantity $F$
only depends on these system parameters.
In other words one may define a so-called \newterm{observable}
$F\clphase$ as a function $\R^{2d} \to \R$, \ie
from a vector in \newterm{phase space} to the measured value.
Clearly the coordinates $q_i$ and momenta $p_i$ are observables as well.
The most important observable is the total energy function or \newterm{Hamiltonian}
\begin{equation}
	H\clphase
	\equiv \underbrace{\frac{1}{2} \sum_{k=1}^d \frac{p_k^2}{m_k}}_{= T(p_1, \ldots, p_d)}
	+ V(q_1, \ldots, q_d),
	\label{eqn:HamiltonianClassical}
\end{equation}
where $m_k$ is the mass of the particle associated with
the degree of freedom $k$,
$T$ is the kinetic energy observable and $V$ the total potential energy observable.
In the formalism of Hamiltonian mechanics, $H$ governs the time evolution
of the system, namely
\begin{align}
	\frac{\D p_k}{\D t} &= - \frac{\partial H}{\partial q_k}, &
	\frac{\D q_k}{\D t} &= \frac{\partial H}{\partial p_k}
	&&\forall k \in \{1, \ldots, d\}.
	\label{eqn:HamiltonianEqnMotion}
\end{align}

These expressions allow to generalise the description of the time evolution
to any other arbitrary observable as well.
To make the connection to quantum mechanics more apparent,
let us introduce for this purpose the so-called \newterm{Poisson bracket}.
It is the skew-symmetric form
\begin{equation}
	\left\{ F, G \right\}_P \equiv \sum_{j=1}^d \left(
	\frac{\partial  F}{\partial q_j} \frac{\partial G}{\partial p_j}
	- \frac{\partial  F}{\partial p_j} \frac{\partial G}{\partial q_j}
	\right).
	\label{eqn:PoissonBracket}
\end{equation}
According to the Liouville equation
\begin{equation}
	\frac{\D F}{\D t} = \frac{\partial F}{\partial t} + \left\{ F, H \right\}_P.
	\label{eqn:PoissonTimeEvolution}
\end{equation}
it relates the Hamiltonian $H$ to the time evolution of any arbitrary observable $F$.
One may also show the relationships
\nomenclature{$\delta_{ij}$}{Kronecker delta. Equal to $0$ if $i \neq j$, else $1$.}
\begin{align}
	\{ q_k, q_l \}_P = \{p_k, p_l\}_P &= 0 & \{ p_k, q_l\}_P
	= - \delta_{kl} &&\forall k,l \in \{1, \ldots, d\}
	\label{eqn:PoissonProperties}
\end{align}
between the principle system observables.

\subsection{Moving to quantum mechanics}
\label{sec:IntroQM}
\defineabbr{QM}{QM\xspace}{Quantum mechanics}
We will now introduce (non-relativistic)
quantum mechanics~(QM) in a rather pragmatic manner,
namely by stating a summary what changes in the \QM formulation
compared to the classical one.
The full mathematical details are not yet stated at this point
and not all terms defined.
The reader should take this overview as a motivation
for a more detailed treatment of the mathematical concepts
further down this chapter.

\nomenclature{$\hilbert$}{Arbitrary Hilbert space,
	typically the Hilbert space of quantum mechanics, which is $L^2(\R^d, \C^s)$
	for $d$ spatial and $s$ spin dimensions.
}
\begin{enumerate}[label=(\alph*)]
	\item Instead of phase space vectors $\clphase \in \R^{2d}$,
		in \QM a particular state of the system is represented
		by functions $\Psi : \R^d \to \C$
		mapping from space coordinates to a complex number.
		These originate from a complex separable Hilbert space $\hilbert$.
	\item A classical observable $F$ is represented by a
		self-adjoint operator $\Op{F}$ on the Hilbert space $\hilbert$ in \QM.
	\item For each classical observable one may construct an equivalent
		corresponding \linebreak
		quantum-mechanical operator.
		For example the observable $q_k$ corresponds%
		\footnote{In fact alternative constructions for the position and momentum
				operator are possible as well. In this work the so-called
				\textit{position representation} is presented.}
		to~\cite{Shankar1994, Mueller2000}
		\[ q_k \longrightarrow \op{x}_k = x_k, \]
		\ie just the multiplication with $x_k$,
		the $k$-th coordinate of the system.
		On the other hand $p_k$ corresponds to~\cite{Shankar1994, Mueller2000}
		\[ p_k \longrightarrow \op{p}_k
			= \frac{\hbar}{\I} \frac{\partial}{\partial x_k}, \]
		an appropriately scaled derivative with respect to $x_k$.
	\item Relationships originating from classical mechanics
		can (usually) be transformed into their \QM analogue.
		For this replace all occurrences of the Poisson bracket
		and the contained classical observables with the commutator
		\begin{equation}
			\comm{\Op{F}}{\Op{G}} = \Op{F} \Op{G} - \Op{G} \Op{F}
			\label{eqn:Operator}
		\end{equation}
		and corresponding operators, \ie~\cite{Shankar1994}
		\[ \{F, G\}_P \longrightarrow \frac{\I}{\hbar} \comm{\Op{F}}{\Op{G}}. \]
	\item The measured values of an observable $F$
		are the eigenvalues $\lambda_k$ of $\Op{F}$ \emph{only}.
	\item Assume that we can find a complete and countable
		set of eigenpairs $\{ (\lambda_\mu, \Psi_\mu) \}_{\mu \in \mathcal{I}}$
		for the operator%
		\footnote{We will see in section \vref{sec:spectral}
		that this is \emph{not} always possible.
		A more general treatment of spectral theory involving
		\textit{spectral projectors} allows to reformulate point (f) for cases
		where such eigenpairs cannot be found.
		See \cite{Helffer2013} for details.}
		$\Op{F}$, where $\mathcal{I}$ is an appropriate index set.
		One may then compute the expectation value of a measurement
		on a normalised%
		\footnote{unit normalised, \ie $\braket{\Phi}{\Phi}_\hilbert = 1$}
		state $\Phi$ as
		\begin{equation}
			\langle \Op{F} \rangle
				= \braket{\Phi}{\Op{F} \, \Phi}_\hilbert
				= \sum_{\mu\in\mathcal{I}} \lambda_\mu \,
							\abs{\braket{\Psi_\mu}{\Phi}_\hilbert}^2,
			\label{eqn:QMExpectation}
		\end{equation}
		where $\braket{\slot}{\slot}_\hilbert$ is the inner product
		of the Hilbert space $\hilbert$.
\end{enumerate}
Now that we discussed the \textit{ad hoc} modification of classical mechanics
in order to yield a theory based on the postulates of \QM,
let us see how one is able to deduce useful results of \QM
from the analogous expressions in classical mechanics.
For example from \eqref{eqn:PoissonProperties} we can immediately
deduce important commutator relations between the position and momentum operators:
\begin{align}
	\comm{\op{x}_k}{\op{x}_l} &= \comm{\op{p}_k}{\op{p}_l} = 0 & \comm{\op{x}_k}{\op{p}_l} = \frac{\hbar}{\I} \delta_{kl}.
	\label{eqn:compatibilityOperator}
\end{align}
Similarly from \eqref{eqn:PoissonTimeEvolution} we obtain
\begin{equation}
	\frac{\D \Op{F}}{\D t} = \frac{\partial \Op{F}}{\partial t} + \frac{\I}{\hbar} \comm{\Op{F}}{\Op{H}},
	\label{eqn:HeisenbergEqnMotion}
\end{equation}
the equation of motion, which governs the time-evolution of the operator $\Op{F}$
in the so-called \newterm{Heisenberg picture} of \QM.
Taking the statistical average over \eqref{eqn:HeisenbergEqnMotion}
results in the Ehrenfest theorem
\begin{equation}
	\frac{\D \langle \Op{F} \rangle}{\D t} = \frac{\partial \langle \Op{F} \rangle}{\partial t} + \frac{1}{\I \hbar} \left\langle \comm{\Op{F}}{\Op{H}} \right\rangle.
	\label{eqn:EhrenfestTheorem}
\end{equation}
This result allows to rationalise the \newterm{correspondence principle}
of classical mechanics and \QM,
which we have developed so far.
Comparing \eqref{eqn:EhrenfestTheorem} and \eqref{eqn:PoissonTimeEvolution}
and keeping in mind that the expectation value $\langle \Op{F} \rangle$
as well as the classical observable $F$ both tell us about the result
of a measurement,
we can deduce that --- on average --- classical mechanics still holds.
Thus we may expect the classical expressions to carry some meaning
in the \QM sense as well.

\subsection{The Schrödinger equation}
Even though the Heisenberg picture developed
in the previous section is descriptive for deducing the analogy between
classical mechanics and \QM,
it is not very suitable for the kinds of problems we will be looking
at in the remainder of this thesis.
More suitable for our needs is the \newterm{Schrödinger picture} of \QM,
which differs in the way it treats time evolution.
In the Heisenberg picture the state function $\Psi$ is time-independent
and the operators evolve.
In the Schrödinger picture it is the other way round, i.e. $\Psi$ may change over time
and the operators are static.
Both pictures are related by a unitary transformation
in the Hilbert space $\hilbert$
governed by the Stone-von Neumann theorem\cite{Neumann1931,Neumann1932,Stone1932}.

The equivalent expression to \eqref{eqn:HeisenbergEqnMotion}
in the Schrödinger picture is the \newterm{time-dependent Schrödinger equation}%
~\cite{Shankar1994,Mueller2000}
\begin{equation}
	\Op{H} \Psi = \I \hbar \frac{\partial}{\partial t} \Psi.
	\label{eqn:TDSE}
\end{equation}
Similar to \eqref{eqn:HeisenbergEqnMotion} the key operator governing the
time-evolution of the system is $\Op{H}$.
By analogy to its classical counterpart $\Op{H}$ is referred to as the
\newterm{\QM Hamiltonian} or just Hamiltonian as well.
In fact many properties of a system may already be deduced by considering the
eigendecomposition of its Hamiltonian $\Op{H}$ alone.
In light of this the ansatz
\begin{equation}
	\Op{H} \Psi_\mu = E_\mu \Psi_\mu
	\label{eqn:TISE}
\end{equation}
for finding the Hamiltonian's eigenpairs
\[ (E_\mu, \Psi_\mu) \in \R \times \hilbert \]
is given the name \newterm{time-independent Schrödinger equation}~(TISE) as well.
\defineabbr{TISE}{TISE\xspace}{Time-independent Schrödinger equation}

\pagebreak
\subsection{The Hamiltonian of the hydrogen-like atom}
\nomenclature{$\Delta$}{Laplace operator.}
\nomenclature{$\Op{T}$}{Kinetic energy operator.}
\nomenclature{$\Op{V}$}{Potential energy operator.}
Employing the correspondence principle it is often very convenient
to construct the \QM Hamiltonian of a system starting
from the classical energy expression.
Let us consider a hydrogen-like system,
where a particle of positive charge $Ze$
is clamped at the origin and surrounded by a single electron.
In Cartesian coordinates the position of the electron can be described
by the vector $\vec{r} \equiv \rtp{(x_1, x_2, x_3)}$ and its momentum by
the vector $\vec{p}$.
The classical kinetic energy and potential energy of such a system
are given by
\begin{align}
	T &= \frac{\vec{p} \cdot \vec{p}}{2 m_e} &
	&\text{and}&
	V &= - \frac{Ze^2}{\norm{\vec{r}}_2} = - \frac{Ze^2}{r}. \\
\intertext{respectively. The appropriate \QM analogues are}
	\Op{T} &= -\frac{\hbar^2}{2 m_e} \Delta &
	&\text{and}&
	\Op{V} &= - \frac{Ze^2}{r}
	\label{eqn:OpHydrogenConventional}
\end{align}
where the \newterm{Laplace operator} in $3$ dimensions
\begin{align}
	\Delta = \sum_{i=1}^3 \frac{\partial^2}{\partial x_i^2}
	\label{eqn:LaplaceOperatorHydrogen}
\end{align}
was introduced. The full Hamiltonian therefore reads
\begin{equation}
	\Op{H} = \Op{T} + \Op{V} = -\frac{\hbar^2}{2 m_e} \Delta - \frac{Ze^2}{r}.
	\label{eqn:OpHydrogenFullUnits}
\end{equation}

The eigenvalues of the hydrogen-like Hamiltonian \eqref{eqn:OpHydrogenFullUnits}
can be determined analytically,
see section \vref{sec:HydrogenAtom} for details.
The lowest eigenvalue is an energy of around $\unit[-2.18\E{-18}]{J}$,
hardly a convenient number.
In fact most energy values in quantum chemistry and physics
are of this order of magnitude.
Similarly many other relevant quantities like the charge of the involved
particles or typical lengths are only very small numbers.
For this reason so-called \newterm{atomic units} are typically employed.
These are generated by a unitary transformation of the Hilbert space,
which effectively yields
\[ \hbar \equiv e \equiv m_e \equiv a_0 \equiv E_h \equiv 1. \]
See table \vref{tab:AtomicUnits} for the values of these quantities
in terms of the usual SI units.
%
Employing this transformation on \eqref{eqn:OpHydrogenFullUnits} gives rise to
\begin{equation}
	\Op{H} = -\frac12 \Delta - \frac{Z}{r}
	\label{eqn:OpHydrogen}
\end{equation}
for the hydrogen-like Hamiltonian in atomic units.
Its lowest energy eigenvalue is $-1/2E_h$,
certainly a more pleasant number.
Additionally the sketched transformation has simplified
the expression of the operator from \eqref{eqn:OpHydrogenFullUnits}
to \eqref{eqn:OpHydrogen},
which is in fact a general observation for the relevant
equations of quantum physics and chemistry,
providing another justification for their use.
From now on we will work exclusively in these units.
A more detailed discussion of atomic units,
approaching the subject from a slightly different angle can be found
in \cite{Quack2011}.

\begin{table}
	\centering
	\begin{tabular}{lllll}
		\toprule
		symbol & name & atomic unit of & value in SI units \\
		\midrule
		$\hbar$ & Planck constant & action & $\unit[1.055\E{-34}]{Js}$ \\
		$e$ & elementary charge & charge & $\unit[1.602\E{-19}]{C}$ \\
		$m_e$ & electron mass & mass & $\unit[9.109\E{-31}]{kg}$ \\
		$a_0$ & Bohr radius & length & $\unit[5.292\E{-11}]{m}$ \\
		$E_h$ & Hartree energy & energy & $\unit[4.360\E{-18}]{J}$ \\
		\bottomrule
	\end{tabular}
	\caption[Atomic units and their relationship to SI units]
		{Atomic units and their relationship to SI units.
			Values taken from \cite{CODATA2014}.}
	\label{tab:AtomicUnits}
\end{table}
