\section{Related quantum-chemical software packages}
\newcommand{\psifour}{\texttt{Psi4}\xspace}
\newcommand{\pyquante}{\texttt{pyQuante}\xspace}
\newcommand{\horton}{\texttt{HORTON}\xspace}
\newcommand{\gpaw}{\texttt{GPAW}\xspace}
\newcommand{\ASE}{\texttt{ASE}\xspace}
\newcommand{\CPtK}{\texttt{CP2K}\xspace}

\label{sec:related}
% Some of the programs mentioned here were discovered using
%    https://en.wikipedia.org/wiki/List_of_quantum_chemistry_and_solid-state_physics_software
%

This section will briefly review the efforts of other program packages,
which similar to \molsturm focus on flexibility or generality.
Our discussion here is not intended to be exhaustive,
much rather we only want to present an overview of the status
in current quantum-chemistry softwares with respect to these aspects.

When it comes to flexibility of a program package
a key ingredient is a versatile interface.
This allows to invoke or extend the methods already available elsewhere.
Recently the scripting language \python has become very popular
for achieving this.
Even even meta-projects like \ASE~\cite{Larsen2017},
which aim at extending existing packages by a common \python front end,
have emerged.
Other packages like \horton~\cite{Verstraelen2017}, \pyscf~\cite{Sun2017},
\pyquante~\cite{PyQuante} and \gpaw~\cite{Mortensen2005,Enkovaara2010} are written
almost exclusively in \python and only employ low-level \ccc or \cpp
code for the computational hot spots to various extents.
Starting from the opposite direction \psifour~\cite{Parrish2017} has
gradually introduced a more and more powerful \python interface on top of
their existing \cpp core over the years.

The popularity of the combination of \fortran or a \ccc-like
language in the core and \python as the high-level interface language
can be understood by considering the publication of \citet{Sun2017}
about the \pyscf package.
They rationalise the choice of \python as follows:
\begin{itemize}
	\item There is no need to learn a particular domain-specific
		input format.
	\item All language elements from \python are immediately
		available to \eg automatise repetitive calculations
		with loops or similar.
	\item The code is easily extensible beyond what is available
		inside \pyscf, for example to facilitate plotting
		or other kinds of analysis.
	\item Computations can be done interactively,
		which is helpful for testing or debugging.
\end{itemize}
We would like to add here,
that \python as a high-productivity, multi-paradigm language
often permits to achieve even complicated tasks with few lines of code,
whilst still staying surprisingly easy to read.
In the context of quantum chemistry
this has the pleasant side effect that a \python script
used for performing calculations and subsequent analysis
is typically not overly lengthy,
but still documents the exact procedure which is followed.
All this comes at pretty much no downside
if \python is combined with
carefully optimised low-level \ccc or \fortran
code in the computational hot spots.
\citet{Sun2017} for example claim that \pyscf is as
fast as any other existing quantum chemistry packages
written solely in \ccc or \fortran.

Another common feature between \pyscf and \psifour
is their modular design inside the package.
They vividly facilitate well-established open standards
like HDF5~\cite{HDF5Manual} or \numpy arrays~\cite{Walt2011}
for data exchange,
such that linking their codes to external projects is easily feasible.
\psifour for example managed to integrate more than 15 external packages
into their framework.
This includes three completely different back ends for the computation of the
required integrals.
In the case of \pyscf it only took us about a day to link
our \molsturm to the FCI algorithms of \pyscf
via an interface based on \numpy.
Nevertheless the numerical requirements of Gaussian-type orbitals
are currently hard-coded inside the optimised
\ccc or \cpp parts of both these projects,
such that extending them by other types of basis functions could still be involved.

With respect to supporting a large range of basis function types,
especially the quantum Monte Carlo packages CASINO~\cite{Needs2010}
and QMCPACK~\cite{Kim2012} should be mentioned.
They both allow to start a quantum Monte Carlo calculation
from discretisations of the trial wavefunction
in terms of GTOs, STOs, plane-waves or numerical orbitals like splines.
Similarly the packages \CPtK~\cite{Hutter2014}, \ASE~\cite{Larsen2017}
and \gpaw can be employed to perform and post-process computations
using more than one type of basis function.
Both \gpaw and \CPtK even allow to perform calculations
employing hybrid basis sets with a mixture of
Gaussian-type orbitals and plane waves.
To the best of our knowledge,
the design of these packages is, however,
very specific to the particular combinations of basis function type
and method they support.
\todoil{I (mfh) would leave this sentence out, since it is a repetition
	of what was said in the previous paragraph and is kind of implicitly clear: \\[1em]
Most notably implementing basis function types or methods with
deviating numerical requirements from the present algorithms
could still be rather involved.
}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "paper"
%%% End:
