\chapter{Numerical approaches for solving the Hartree-Fock problem}
\chaptermark{Numerical approaches for solving \HF}
\label{ch:NumSolveHF}
\chapquote{%
I believe there is no philosophical high-road in science,
with epistemological signposts.
No, we are in a jungle and find our way by trial and error,
building our road behind us as we proceed.
}{Max Born~(1882--1970)}
\newcommand{\kexp}{\ensuremath k_\text{exp}}

This chapter is devoted to an in-depth discussion of
numerical approaches for solving the \HF problem
both when it comes to the basis function type used for the discretisation
and the algorithms for solving the discretised problem.
We will discuss how different basis function types
lead to numerical problems of vastly different structure
and how therefore not every algorithmic ansatz works
for every type of basis function.

In section \ref{sec:DiscreteHF} we noted that there are
roughly three ways to view the discretised \HF problem.
One way would be to think of it as a minimisation of the energy
with respect to the orbital coefficients,
another as a minimisation with respect to the density matrix
and yet a third as a non-linear eigenproblem,
which needs to solved self-consistently.
Our discussion here will generally take the third viewpoint
and only switch to the others when this aids our argument.
Furthermore we will implicitly assume a real-valued
\UHF ansatz in this chapter.
The adaption of the presented results to
\RHF or \ROHF is usually straightforward%
\footnote{
To go from \UHF to \RHF one just needs to consider both blocks of the relevant
Fock, coefficient, and density matrices to be equivalent.
Going from \UHF to \ROHF only amounts to replacing the \UHF Fock matrix
by the appropriately constructed \ROHF Fock matrix before performing
the diagonalisation for getting the new coefficients.
}.

\input{4_solving_hf/scf_overview.tex}
\input{4_solving_hf/guess.tex}
\input{4_solving_hf/basis_functions.tex}
\input{4_solving_hf/scf_algos.tex}

\section{Basis-type independence of SCF algorithms}
% Show schematic with Fock update - coefficient update
%     in red and blue as in paper
% Refer to appropriate equations from more detailed considerations in SCF
% algorithms chapter.
% TODO This could fit into the introductory part of the SCF algorithm
% section or in the conclusion part of it as well


\todo[inline,caption={}]{
	\begin{itemize}
		\item Make sure it comes across that SCF has very similar structure for each basis type
		\item SCF algorithm may be formulated independent of basis type
		\item DFT very similar
		\item[$\Rightarrow$] Basis-function independent quantum chemical modelling
		\item Maybe move to flexible qchem program chapter
	\end{itemize}
}
