\chapter*{Acknowledgements} \phantomsection
\markboth{\leftheadstyle{Acknowledgements}}{}
\addcontentsline{toc}{chapter}{Acknowledgements}
\todoil{Draft version}

First and foremost I would like to thank
my supervisor Prof. Andreas Dreuw,
my co-supervisor Prof. Guido Kanschat
as well as my advisor Dr. James Avery.
They provided me with a constant source of advice, support
and helpful comments throughout the recent years.
Andreas, thanks for giving me the chance to work on such
a highly interdisciplinary project
and for allowing me the scientific freedom
to change the original topic into what the thesis has become.
Guido, thanks for the enlightening discussions about numerics
as well as the occasional criticism to make me invest a second thought.
James, thanks for introducing me to Coulomb-Sturmians,
for the great time I had whenever I visited you in Copenhagen
and of course for showing me Heisenberg's bath tub.

Next I wish to express my gratitude to Dr. Michael Wormit,
with whom I was able to work,
albeit for too short a period of time.
It brings me a consoling thought that some of his
ideas managed to yield fruit in this thesis.

% -----

An aspect of quantum chemistry,
which has always fascinated me,
is the interdisciplinarity of the field.
It is absolutely astonishing
to find that the current status could only be achieved
by a combination developments guided by chemical as well as physical intuition,
coupled with a careful study of the beautiful and complex
underlying mathematical structures
as well as many dirty tricks of high-performance computing.
In this respect I am very grateful
to all the impulses and ideas from
every scientific field and research topic I was exposed to.
I wish to express my thanks for all the discussions with
\todo{randomise and commas}
Prof. Eric Cancès
Prof. Reinhold Schneider
Dr. Denis Davydov
Dr. Katharina Kormann
%
Adrian Dempwolff
Dr. Tim Stauch
Manuel Hodecker
Maximilian Scheurer
%
Dr. Klaus Birkelund
Dr. Mads Kristensen
%
Henrik Larsson
Dr. Jenny Wagner
Lucas Fabian Hackl
Jan Janßen
Jan-Christoph Peters
Fabian Klein
and everyone I forgot to mention in this list.

% -----

In the same respect I would like to acknowledge the open and curious
people I met at all hacker events I've been to so far
with many nights of talking
about science, society, ethics, programming and just general nerdiness.
A special thanks goes to 
\todo{randomise and commas}
cherti
hauro
janx2
kungi
rami
supaake
and all the other guys from the NoName e.V.

% -----

I gratefully acknowledge
the University of Kopenhagen,
KTH Royal Institute of Technology,
Mathematical Research Institute of Oberwolfach
as well as above all
Heidelberg University
for hosting me.
Further I thank the Heidelberg Graduate School
for the Mathematical and Computational Methods for the Sciences
for providing me with financial support
and for the opportunity to take action myself
in the form of taking part in organising conferences
or giving block courses to fellow students.

I wish to express my thanks
to the past and present members of the Dreuw group,
for the relaxed atmosphere and for being more than just colleagues.
Thanks for all the fun
at our retreats in Kleinwalsertal, our trips to Wurstmarkt,
hiking tours in Grünstadt,
all the beers and wines after work
and for the running gags about stuffed animals,
singers and writers,
whose should better not be named in my thesis.

In the same way I wish to thank
the group of Prof. Dr. Brian Vinter
at the Niels Bohr Institute
for the open and welcoming atmosphere I was presented with.
There is no question I immediately felt part
and truely enjoyed the time I spent with you,
let it be the Monday morning breakfast meetings
or the beer tastings.
I am already looking forward to coming back for another visit!

I gratefully acknowledge the computational time I was provided
at the computing facilities of the Vinter group,
as well as all the excellent support from Jonas Bardino
with respect to getting my code to run.
In the same way I thank Manfred Trunk
for having an open door for all requests related
to our cluster in Heidelberg
and for providing almost instant answers for all emails I sent.
Thanks to Ellen Vogel for taking care of the bureaucracy.

% -----

With respect to the thesis I thank
\todo{randomise}
Fabian Faulstich
Dr. Jenny Wagner
Adrian Dempwolff
Manuel Hodecker
Maximilian Scheurer
Henrik Larsson
Marvin Hoffmann 
Dr. James Avery
% Reena Sen
% Mikael Scott
% Dr. Jim Dengler
% Carine Dengler
who provided helpful comments and feedback
by looking at drafts and correcting errors meticulously.
Furthermore I wish to express my thanks to the many people developing
the software, without which this thesis could not have been made,
namely
\texttt{bash},
\texttt{clang},
\texttt{gcc},
\texttt{i3},
\LaTeX,
\numpy,
\python,
\TeX,
\texttt{vim}
and all the uncountably many scripts and utilities
I use everyday without being grateful enough for their work.
Especially I want to acknowledge all the people
involved with providing the best 
GNU/LinuX distribution in the world,
namely Debian.

% -----

Thanks to my old friends from the days in Grünstadt,
\todo{randomise}
Jan-Christoph Peters,
Peter Schwalb,
Alexandra Schulte
and
Jan Janßen
which have accompanied me for many years
on hiking trips or found the time
for a spontaneous visit.
I know we have met fare too rarely in the recent months.

Further I would like to thank Henrik Larsson
for a lasting friendship,
which has lead to many weekends of fruitful
scientific debate and too often new insights into my own subject.
I like that we do not always agree
and that the only way to convince you is scientific evidence.
Thanks for accompanying me again
to lecture a course at the Deutsche Schülerakedemie this summer.

% TODO  sounds stupid
The same hold true for all of my other friends.
I am really greateful for every minute we spent and will spend.

% ----

E große Dank sei ach
mei ganz Familie vun de Palz.
Ganz bsonders will ich danke Sven, Oliver,
Michael, Markus un Jun,
als ach mene Unkel Klaus un Hans
un meine Tante Gerda, Edith und Änne.
Mit eich ebbes zu unnernemme oder efach blos zu babble
des war oft e Quell vun Kraft fer die mehr anstrende Daage.
Vielmols merci sag ich dem Sonja und dem Jim
fer die viel schene Zeit in de letzte Johre.
Danewe dank ich meine Eltre Ingrid un Ortwin Herbst,
bei dene efache Worte eigentich net genuch sind
um des im entferndeschde zu fasse.

% ----

Finally and above all I thank my beloved fianc\'ee Carine
for every minute we have been, are and will be together.
